Human capital mediates natural selection in contemporary humans

### 2022-05-03 response-to-reviewers-BE.Rmd

Thank you very much for the chance to revise our manuscript. We would also like to thank both reviewers for their comments. Below we deal with the points they raised.

We've also made some other changes:

-   We moved the results on causality of polygenic scores to the appendix. We thought they were a distraction from the key point of Section 4, which is to test our theory.

-   Code is now publicly available at <https://github.com/hughjonesd/why-natural-selection>.

@@

We have changed vocabulary accordingly, using "stabilizing" and "disruptive" throughout.

> p\. 10: I am interested in the functional form chosen for u(y). This has sensible properties for sigma \< 1 near 1, including being increasing in y for y somewhat greater than zero and being concave. Intuitively, this means that utility increases with income (net of costs of childrearing) but less so as income increases. Is there any particular reason for this functional form as opposed to others that also satisfy these properties? Also, is there any empirical evidence that "utility" obeys this relationship?

The functional form is a standard one in economics. It has the specific property that relative risk aversion is constant. That is, if you have CRRA utility and £100, and you are indifferent to a 50-50 bet between winning £30 and losing £20, then if you had £1000 you would be indifferent between a 50-50 to win £300 or lose £200.

The literature on estimating utility functions is vast (see e.g. Chuang and Schechter 2015). One issue is that of context. CRRA may make sense for risky financial decisions. Here, as Reviewer 2 points out, the "curvature" of the utility function is used not to explain attitudes to risk, but to trade off income against children at different income levels. This author wishes there was a single utility function that worked to explain all human behaviour across contexts... in reality, economists tend to use "what works". We use the CRRA form because it is tractable; we expect that using other forms, intermediate levels of risk aversion would still generate our empirical results, but the maths would be uglier. For a useful discussion of microeconomic theories of fertility, including functional forms, see Jones et al. (2008) and more recently Doepke et al. (2022) which uses a very similar form to ours to capture birth timing.

@@

> \> More generally, I think the authors can improve their empirical analysis of human capital as a mediator of natural selection (the above comments should help with this).

We welcome any further suggestions.

> \
> 3. To adjust for ascertainment bias, the authors adjust their estimates using three weighting schemes. Why not combine all the variables from the three schemes and add other observables that could impact ascertainment (including also sex, proximity to a recruitment center, number of individuals in the household, and household income)? If the authors' goal is to adjust their estimates for ascertainment in the UKB, they should use all relevant observables to create UKB weights so that their estimates more closely resemble those they'd obtain in a more representative sample. This is particularly important here since we're interested in selection, and estimates can be severely biased in a selected sample like the UKB.\
> \
> Also, instead of presenting a small subset of the results with various weighting schemes in Figure 2, it'd be preferable for the authors to pick what they think is the best weighting scheme and then present all results with that scheme. Then, they can briefly discuss the robustness of their results to various weighting schemes.

We agree with this. Combining all the variables is hard: age at first live birth is only available for females, and some of the data comes from different sources. In the new version, we use weights kindly provided to us by Sjoerd van Alten and coauthors. Their working paper is specifically dedicated to weighting UK Biobank, so it seems good to use the "industry standard" solution rather than rolling our own. We now use these weights throughout, but detail results from alternative weighting schemes in the appendix.

@@

> 4\. On p. 4, line 13, the authors write that "Beta is the expected polygenic score among children of the sample", with a footnote explaining the reason. However, since the authors use N=number of children rather than relative number of children (=number of children/mean number of children), I don't think that's correct. See Beauchamp PNAS 2016. More generally, why not use relative lifetime reproductive success (rLRS) instead of number of children as the dependent variable in equation 1?

Oops, yes. We hope the corrected footnote is now right.

We now use RLRS as the dependent variable for both respondents and parents. Beyond the obvious change in absolute effect sizes, results are not much changed, except there is now slightly more evidence for larger selection effects over time: 8 polygenic scores show a significant increase in effect size on respondents' RLRS when we median-split them by birth year. Probably this was masked when we used absolute fertility, by the fall in fertility over time. In turn, this makes us less sure that we can rule out that the welfare state is a relevant explanation for natural selection, so we have slightly qualified our comments on this topic.

@@

Abdellaoui, A., Hugh-Jones, D., Yengo, L., Kemper, K.E., Nivard, M.G., Veul, L., Holtz, Y., Zietsch, B.P., Frayling, T.M., Wray, N.R. and Yang, J., 2019. Genetic correlates of social stratification in Great Britain. *Nature human behaviour*, *3*(12), pp.1332-1342.

Bergsvik, J., Fauske, A. and Hart, R.K., 2021. Can Policies Stall the Fertility Fall? A Systematic Review of the (Quasi‐) Experimental Literature. *Population and Development Review*, *47*(4), pp.913-964.

Chuang, Y. and Schechter, L., 2015. Stability of experimental and survey measures of risk, time, and social preferences: A review and some new results. *Journal of development economics*, *117*, pp.151-170.

Doepke, M., Hannusch, A., Kindermann, F. and Tertilt, M., 2022. *The economics of fertility: A new era* (No. w29948). National Bureau of Economic Research.

### 2022-05-03 why-natural-selection.Rmd
Fix a bug in run_regs_fml(); numerous tweaks to text

drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_all)
drake::loadd(res_sex)

res_wt_combined <- bind_rows(
  None                = res_all %>% 
                          filter(
                            reg.type == "controlled", 
                            dep.var == "RLRS"
                          ) %>% 
                          select(-reg.type, -dep.var, -score_name),
  Geographical        = res_wt_msoa_weights,
  `Age/Qualification` = res_wt_age_qual_weights,
  `Age/Qual/AFLB`     = res_wt_flb_weights,
  .id = "Weights"
)

@@

To correct for ascertainment bias, we weight participants using population data.
We try three alternative weighting schemes: (1) "Geographical" weighting, by
geography, age and presence/absence of a partner; (2) age and highest
educational qualification; and (3) for women only, age, highest qualification,
and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
selection effects among the entire sample, estimated with the three weighting
schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
scores are significant at $p < 0.05/33$ under all four weighting schemes, with
`r res_sig_any` scores reaching significance under at least one scheme.[^balance-diverse-def]
Weighting makes a large difference to estimates: mean effect sizes across all
polygenic scores are increased by a factor of `r mean_rel$Geographical`
(geographical weighting), `r mean_rel[["Age/Qualification"]]`
(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
Estimates might be further affected by weighting on other demographic
variables.[^weighting]

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their
correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots
selection effects on each polygenic score against that score's correlation with
two measures of human capital: earnings in a respondent's first job, and
educational attainment. The relationships are strongly negative. Thus, human
capital appears to be relevant to natural selection. The negative relationship
suggests that substitution effects dominate income effects, which fits the known
negative association between income and fertility
[@becker1960economic;@jones2006economic]. The correlations reverse when we
control for age at first live birth, suggesting that within AFLB categories, the
income effect dominates. (See also the Supplementary Animations for the
uncorrected effects per AFLB group, which show the correlation reversing after
age 22.)

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, the majority of effect sizes appear unchanged
over a large span of twentieth-century history (Appendix Table
\@ref(tab:tbl-change-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. Future
work could test this theory more explicitly. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital. A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our parameter $\sigma$, which can be interpreted as a measure of risk aversion
over income. However, a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced, or whether genetic
variants underlying other phenotypes will show a similar pattern to those
studied here. In addition, effects of polygenic scores may be inflated in
population-based samples, because of indirect genetic effects, gene-environment
correlations, and/or assortative mating [@lee2018gene;
@selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935], although we do
not expect that this should impact their association with number of offspring,
or the resulting changes in allele frequencies. In short, it is probably too
early to tell whether modern natural selection has a substantively important
effect on population means of phenotypes under selection. Nevertheless, we note
that selection effects on our measured polygenic scores are still relatively
small, even after reweighting to account for ascertainment bias.

@@

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects. Respondents weighted by Age/Qualification.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )

@@

Existing evidence on human natural selection has led some to "biocosmic
pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
natural selection's effects are outweighed by environmental improvements, such
as those underlying the Flynn effect [@flynn1987massive]. The evidence
here may add some nuance to this debate. Patterns of natural selection have been
relatively consistent across the past two generations, but they are not the
outcome of a universal, society-wide phenomenon. Instead they result from 
opposing forces, operating in different parts of society and pulling in
different directions.

Any model of fertility is implicitly a model of natural selection, but so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of natural selection in
modern societies. Economics possesses a range of theoretical models on the
effects of skills, education and income [see @hotz1997economics;
@lundberg2007american]. One perennial problem is how to test these theories in a
world where education, labour and marriage markets all interact. Genetic data,
such as polygenic scores, could help to pin down the direction of causality,
for example via Mendelian randomization [@davey2003mendelian].
Conversely, theory and empirical results from economics can shine a light on the
mechanisms behind natural selection, and thereby on the nature of individual
differences in complex traits and disease risk.

@@

Population data for weighting is taken from the 2011 UK Census and the 2006
General Household Survey (GHS). Weighting for Age/Qualification and 
Age/Qualification/AFLB weights was done using marginal totals from a linear
model, using the `calibrate()` function in
the R "survey" package [@lumley2020survey]. Geographical weighting was done with
iterative post-stratification using the `rake()` function, on Census Middle Layer
Super Output Areas, sex and presence/absence of a partner.

@@

\FloatBarrier
\clearpage
## Weighted regressions

@@

rel_sizes %>% 
      arrange(desc(Geographical)) %>% 
      bind_rows(rel_sizes_summary) %>% 
      rename(PGS = term) %>% 
      as_hux() %>% 
      insert_row("", "Weighting", fill = "") %>% 
      set_caption(
        "Weighted effect sizes as a proportion of unweighted effect sizes."
      ) %>% 
      merge_cells(1, 2:4) %>% 
      set_align(1, 2, "centre") %>% 
      set_header_rows(1, TRUE) %>% 
      theme_compact() %>% 
      set_number_format(-1, -1, 2) %>% 
      style_headers(bold = TRUE) %>% 
      set_bottom_border(1, 2:4) %>% 
      set_bottom_border(2, everywhere) %>% 
      set_italic(final(2), 1) %>% 
      set_font_size(8) %>% 
      set_width(0.7) %>% # ensures line wrapping below
      set_col_width(c(.34, .22, .22, .22)) %>% 
      add_footnote(paste0(
        "Only consistently-signed and significant (when unweighted) ",
        "estimates are shown. ",
        "Age/Qual/AFLB as a proportion of unweighted regressions ",
        "including females only."),
        top_padding  = 0,
        left_padding = 0
      )

@@

Stabilizing selection reduces variance in the trait under selection, while
disruptive selection increases variance. To check for these, we rerun equation
\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
age/qualification weights. Scores for hip circumference show significant
stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
The EA2 score for educational attainment shows significant disruptive selection
($p < 0.05/33$, positive coefficient), which reduces the strength of selection
against educational attainment at very high levels of the PGS. (The quadratic on
the EA3 score has a similar coefficient but is not significant at $p < 0.05/33$.)
Figure \@ref(fig:plot-purifying) plots predicted number of children
against polygenic score from these regressions.

We also checked for stabilizing selection in the parents' generation, using
age/qualification weights multiplied by the inverse of number of siblings.
Scores for EA2 and EA3 show significant disruptive selection ($p < 0.05/33$, 
positive coefficient on quadratic). Other scores including hip circumference 
were not significant.

```{r plot-purifying, fig.align = "center", fig.cap = "Stabilizing/disruptive selection: predicted number of children by polygenic score.", fig.height = 3}

@@

stopifnot(sig == c("EA2_noUKB", "hip_combined"))

@@

op <- par(mfrow = c(1, 2))
for (score in sig) {
  coef <- sig_all_terms %>% filter(score_name == score) %>% pull(estimate)
  curve(
         coef[1] + coef[2] * x + coef[3] * x^2,
         xlim = c(-4, 4), 
         xlab = pretty_names(score),
         ylab = "Predicted RLRS"
       )
}
par(op)

@@

Results in Figure \@ref(fig:plot-income-educ-level) could be
explained by age, if older respondents have lower income and are less educated,
and also show more natural selection on polygenic scores. However, when we rerun
the regressions, interacting the polygenic score with income category and also
with a quadratic in age, the interaction with income remains significant at
$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
if we interact the PGS with age of leaving full time education and a quadratic
in age, the interaction with age leaving full time education remains significant
at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.

@@

```{r calc-res-sibs}
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_children_comparison)

cmp <- left_join(res_wt_age_qual_weights, res_children_comparison, by = "term")

@@

The UK Biobank data contains information on respondents' number of siblings
(including them), i.e. their parents' number of children. Since respondents'
polygenic scores are equal in expectation to the mean scores of their parents,
we can use this to look at selection effects in the parents' generation. We
estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
variable.[^parent-RLRS] The parents' generation has an additional source of ascertainment
bias: sampling parents of respondents overweights parents who have many
children. For instance, parents of three children will have, on average, three
times more children represented in UK Biobank than parents of one child. Parents
of no children will by definition not be represented. To compensate, we reweight
our preferred weightings (Age/Qualification) by the inverse of *number of
siblings*.

@@

```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents. Age/qualification weights. Parental generation weights multiplied by 1/number of siblings. Respondents' regression excludes childless respondents."}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

@@

stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 32)
stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 9)
stopifnot(sum(summary_period$Change_Respondents == "Change sign") == 1)
stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 23)
```

@@

As an alternative approach, we run regressions interacting polygenic scores with
birth year, median split at 1950 ("early born" versus "late born"). We use both
respondents' RLRS and parents' RLRS as a dependent variable. We weight
using age/qualification cells, and further adjust for selection in the parents'
generation (see above).

Table \@ref(tab:tbl-change-by-period) 
summarizes the results. We report the number of scores showing significant
changes over time (i.e. a significant interaction between polygenic score and
the "late born" dummy): either a significant change in sign, a significant
increase in size, or a significant decrease in size. There is little evidence
for changes in selection effects within the parents' generation, with just one
score showing a significant decrease in size. In the respondents' generation,
effect sizes were significantly larger in absolute size among the later-born for
nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
Disorder. These changes are inconsistent with the intergenerational change,
where estimated effect sizes were larger among the earlier, parents' generation.
One score, for alzheimer's disease, showed a significant change in sign, from
negative to positive effects on fertility.

Overall, there is some evidence for an increase in the strength of selection in
recent history. The clearest result is that the pattern of relative effect sizes
across scores is broadly consistent over time.

@@

tbl_both %>% 
            rename(
              Change              = Change_Parents, 
              `Parents' RLRS`     = n.x, 
              `Respondents' RLRS` = n.y
            ) %>% 
            as_hux() %>%
            set_bold(1, everywhere) %>% 
            set_bottom_border(1, everywhere) %>% 
            add_footnote(glue("Significance is measured at p < 0.05/{n_tests}."), 
              border = 0.4, font_size = 8) %>% 
            set_top_padding(1) %>% 
            set_bottom_padding(1) %>%
            set_left_padding(everywhere, 1, 0) %>% 
            set_right_padding(everywhere, 3, 0) %>% 
            set_caption("Numbers of polygenic scores showing changes in selection effects between early and late born. 
      Weights for respondents: Age/Qualification. Weights for parents: Age/Qualification times 1/number of siblings.") %>% 
            set_caption_width(1)

@@

```{r plot-siblings-townsend, fig.cap = "Selection effects (parents' RLRS) by Townsend deprivation quintile of birth area. Higher = more deprived. Estimates weighted by 1/number of siblings.", fig.align = "center"}

>>>>

drake::loadd(res_wt_flb_weights)
drake::loadd(res_wt_msoa_weights)
drake::loadd(res_wt_age_qual_weights)
drake::loadd(res_wt_van_alten_weights)
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_unweighted)
drake::loadd(res_sex)

res_wt_combined <- bind_rows(
  None                = res_unweighted %>% 
                          select(-score_name) %>% 
                          filter(term != "(Intercept)"),
  Main                = res_wt_van_alten_weights,
  Geographical        = res_wt_msoa_weights,
  `Age/Qualification` = res_wt_age_qual_weights,
  `Age/Qual/AFLB`     = res_wt_flb_weights,
  .id = "Weights"
)

@@

Figure \@ref(fig:plot-res-wt) plots selection effects among the entire
sample.[^balance-diverse-def] To correct for ascertainment bias, we weight
participants using the model of @van2021effects. Weighting makes a
large difference to estimates, increasing effect sizes by a mean of 
`r percent(mean_rel$Main - 1)`.[^weighting]

@@

res_wt_combined %>% 
      filter(Weights %in% c("Main", "None")) %>% 
      mutate(
        Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "Unweighted"        = "grey15", 
        "Weighted"          = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "Unweighted" = "square filled", 
        "Weighted"   = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

We test this explanation in two ways. First, the economic theory predicts that
genetic variants will be selected for (or against) in proportion to their
correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots
selection effects on each polygenic score against that score's correlation with
two measures of human capital: earnings in a respondent's first job, and
educational attainment. The relationships are strongly negative. Thus, human
capital appears to be relevant to natural selection. The negative relationship
suggests that substitution effects dominate income effects, which fits the known
negative association between income and fertility
[@becker1960economic;@jones2006economic]. The correlations reverse when we
control for age at first live birth, suggesting that within AFLB categories, the
income effect dominates.

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, the majority of effect sizes appear unchanged
over a large span of twentieth-century history (Appendix Table
\@ref(tab:tbl-change-by-period)), during
which government spending on child-related benefits varied considerably
[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
affect fertility [@gauthier2007impact]. Future
work could test this theory more explicitly. A second alternative theory is that
polygenic scores correlate with the motivation to have children, i.e. parameter
$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
selection effects are smaller at higher incomes and education levels. Indeed,
in the model, the effect of $a$ on fertility becomes stronger at higher levels
of human capital. A third alternative is that traits under selection are linked
to externalizing behaviour and risk-seeking. This might be partially captured by
our parameter $\sigma$, which can be interpreted as a measure of risk aversion
over income. However, a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. This theory is less good,
though, at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test the theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
But this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

We also do not know how estimated effect sizes of natural selection will change
as more accurate measures of genetic variation are produced, or whether genetic
variants underlying other phenotypes will show a similar pattern to those
studied here. In addition, effects of polygenic scores may be inflated in
population-based samples, because of indirect genetic effects, gene-environment
correlations, and/or assortative mating [@lee2018gene;
@selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935], although we do
not expect that this should impact their association with number of offspring,
or the resulting changes in allele frequencies. Lastly, although effects on
our measured polygenic scores are small, even after weighting, disadvantages which
are individually small may cumulate and/or interact to create larger effects. 
Overall, it is probably too early to tell whether modern natural selection has a 
substantively important effect on population means.

@@

```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group. Blue is actual. Grey is hypothetical in the absence of selection effects.", fig.width = 4}
income_EA3 <- famhist %>% 
      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
      mutate(
        `Household income` = factor(income_cat, 
                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
                                      "> £100K"))
      ) %>% 
      group_by(`Household income`) %>% 
      summarize(
        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
                                              na.rm = TRUE)
      ) %>% 
      tidyr::pivot_longer(-`Household income`) %>% 
      rename(
        `Children's mean EA3` = value
      )

@@

Existing evidence on human natural selection has led some to "biocosmic
pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
natural selection's effects are outweighed by environmental improvements, such
as those underlying the Flynn effect [@flynn1987massive]. The evidence
here may add some nuance to this debate. Patterns of natural selection have been
relatively consistent across the past two generations, but they are not the
outcome of a single, society-wide phenomenon. Instead they result from 
opposing forces, operating in different parts of society and pulling in
different directions.

Any model of fertility is implicitly a model of natural selection, but so far,
the economic and human genetics literatures have developed in parallel.
Integrating the two could deepen our understanding of natural selection in
modern societies. Economics possesses a range of theoretical models on the
effects of skills, education and income [see @hotz1997economics;
@lundberg2007american]. One perennial problem is how to test these theories in a
world where education, labour and marriage markets all interact. Genetic data,
such as polygenic scores, could help to pin down the direction of causality,
for example via Mendelian randomization [@davey2003mendelian].
Conversely, economic theories and empirical results can shine a light on the
mechanisms behind natural selection, and thereby on the nature of individual
differences in complex traits and disease risk.

@@

Weighting data was kindly provided by @van2021effects.

@@

\FloatBarrier
\clearpage
## Alternative weighting schemes

We report results for 3 alternative weighting schemes: weighting by
age/qualification; geographical (weighting by Middle Super Output Area); and for
women only, age, qualification and age at first live birth. Population data for
weighting is taken from the 2011 UK Census and the 2006 General Household Survey
(GHS). Weighting for Age/Qualification and Age/Qualification/AFLB weights was
done using marginal totals from a linear model, using the `calibrate()` function
in the R "survey" package [@lumley2020survey]. Geographical weighting was done
with iterative post-stratification using the `rake()` function, on Census Middle
Layer Super Output Areas, sex and presence/absence of a partner.

@@

rel_sizes %>% 
      arrange(desc(Main)) %>% 
      bind_rows(rel_sizes_summary) %>% 
      rename(PGS = term) %>% 
      as_hux() %>% 
      insert_row("", "Weighting", fill = "") %>% 
      set_caption(
        "Weighted effect sizes as a proportion of unweighted effect sizes."
      ) %>% 
      merge_cells(1, 2:4) %>% 
      set_align(1, 2, "centre") %>% 
      set_header_rows(1, TRUE) %>% 
      theme_compact() %>% 
      set_number_format(-1, -1, 2) %>% 
      style_headers(bold = TRUE) %>% 
      set_bottom_border(1, -1) %>% 
      set_bottom_border(2, everywhere) %>% 
      set_italic(final(2), 1) %>% 
      set_font_size(8) %>% 
      set_width(0.8) %>% # ensures line wrapping below
      set_col_width(c(.28, .18, .18, .18, .18)) %>% 
      add_footnote(paste0(
        "Only consistently-signed and significant (when unweighted) ",
        "estimates are shown. ",
        "Age/Qual/AFLB as a proportion of unweighted regressions ",
        "including females only."),
        top_padding  = 0,
        left_padding = 0
      )

@@

Stabilizing selection reduces variance in the trait under selection, while
disruptive selection increases variance. To check for these, we rerun equation
\@ref(eq:regression), adding a quadratic term in $PGS_i$. Scores for hip
circumference show significant stabilizing selection ($p < 0.05/33$, negative
coefficient on quadratic term). The EA2 score for educational attainment shows
significant disruptive selection ($p < 0.05/33$, positive coefficient), which
reduces the strength of selection against educational attainment at very high
levels of the PGS. (The quadratic on the EA3 score has a similar coefficient but
is not significant at $p < 0.05/33$.) Figure \@ref(fig:plot-purifying) plots
predicted number of children against polygenic score from these regressions.

We also checked for stabilizing selection in the parents' generation, using
weights multiplied by the inverse of number of siblings. Scores for EA2 and EA3
show significant disruptive selection ($p < 0.05/33$, positive coefficient on
quadratic). Other scores including hip circumference were not significant.

```{r plot-purifying, fig.align = "center", fig.cap = "Stabilizing/disruptive selection: predicted number of children by polygenic score."}

@@

stopifnot(sig == c("ADHD_2017", "bmi_combined", "EA2_noUKB", 
                     "hip_combined", "wc_combined"))

@@

op <- par(mfrow = c(2, 3))

for (score in sig) {
  coef <- sig_all_terms %>% filter(score_name == score) %>% pull(estimate)
  curve(
         coef[1] + coef[2] * x + coef[3] * x^2,
         xlim = c(-4, 4), 
         xlab = pretty_names(score),
         ylab = "Predicted RLRS"
       )
}
par(op)

@@

Results in Figure \@ref(fig:plot-income-educ-level) could be
explained by age, if older respondents have lower income and are less educated,
and also show more natural selection on polygenic scores. However, when we rerun
the regressions, interacting the polygenic score with income category and also
with a quadratic in age, the interaction with income remains significant at
$p$ < 0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
if we interact the PGS with age of leaving full time education and a quadratic
in age, the interaction with age leaving full time education remains significant
at $p$ < 0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.

@@

```{r calc-res-sibs}
drake::loadd(res_wt_van_alten_weights)
drake::loadd(res_children_comparison)

cmp <- left_join(res_wt_van_alten_weights, res_children_comparison, by = "term")

@@

The UK Biobank data contains information on respondents' number of siblings
(including them), i.e. their parents' number of children. Since respondents'
polygenic scores are equal in expectation to the mean scores of their parents,
we can use this to look at selection effects in the parents' generation. We
estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
variable.[^parent-RLRS] The parents' generation has an additional source of ascertainment
bias: sampling parents of respondents overweights parents who have many
children. For instance, parents of three children will have, on average, three
times more children represented in UK Biobank than parents of one child. Parents
of no children will by definition not be represented. To compensate, we multiply
our weights by the inverse of *number of siblings*.

@@

```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents. Parental generation weights multiplied by 1/number of siblings. Respondents' regression excludes childless respondents."}
drake::loadd(res_sibs_parent_weights)
drake::loadd(res_children_comparison)

@@

stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 32)
stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 8)
stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 25)
```

@@

As an alternative approach, we run regressions interacting polygenic scores with
birth year, median split at 1950 ("early born" versus "late born"). We use both
respondents' RLRS and parents' RLRS as a dependent variable. We use our standard
weights, and further adjust for selection in the parents' generation (see above).

Table \@ref(tab:tbl-change-by-period) 
summarizes the results. We report the number of scores showing significant
changes over time (i.e. a significant interaction between polygenic score and
the "late born" dummy): either a significant change in sign, a significant
increase in size, or a significant decrease in size. There is little evidence
for changes in selection effects within the parents' generation, with just one
score showing a significant decrease in size. In the respondents' generation,
effect sizes were significantly larger in absolute size among the later-born for
eight polygenic scores: ADHD, age at menopause, cognitive
ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
Disorder. These changes are inconsistent with the intergenerational change,
where estimated effect sizes were larger among the earlier, parents' generation.

Overall, while there is some suggestive evidence for an increase in the strength
of selection in recent history, the clearest result is that the pattern of
relative effect sizes across scores is broadly consistent over time.

@@

tbl_both %>% 
            rename(
              Change              = Change_Parents, 
              `Parents' RLRS`     = n.x, 
              `Respondents' RLRS` = n.y
            ) %>% 
            as_hux() %>%
            set_bold(1, everywhere) %>% 
            set_bottom_border(1, everywhere) %>% 
            add_footnote(glue("Significance is measured at p < 0.05/{n_tests}."), 
              border = 0.4, font_size = 8) %>% 
            set_top_padding(1) %>% 
            set_bottom_padding(1) %>%
            set_left_padding(everywhere, 1, 0) %>% 
            set_right_padding(everywhere, 3, 0) %>% 
            set_caption("Numbers of polygenic scores showing changes in selection effects between early and late born. Parental generation weights multiplied by 1/number of siblings.") %>% 
            set_caption_width(1)

@@

```{r plot-siblings-townsend, fig.cap = "Selection effects (parents' RLRS) by Townsend deprivation quintile of birth area. Higher = more deprived. Weights multiplied by 1/number of siblings.", fig.align = "center"}
